\section{Implementation} \label{sec:implementation}

This section presents some interesting parts of the implementation of
\texttt{klust}. The documentation for \texttt{klust} is available at the GitHub
website, cf.  section \ref{sec:introduction}, and it is printed when running
the program with no arguments.


\subsection{Compressing sequence data}

Since the sequence data to be clustered can be very large, efficient memory
usage is essential. The implementation of \textsc{K-Clust} reads all the
sequences into memory, while simultaneously transforming it into a lower memory
representation which takes up a quarter of the space of a simple 8-bit
character representation. This is done by representing each nucleotide by just
2 bits.

Currently all the sequence data is stored in memory until the clustering is
finished and the results are written to disk. This strategy eases memory
management, since all sequences can stay in the same place throughout the
clustering and references can be passed around without risking that some
sequence is deallocated too early by some other part of the program.

The \texttt{klust} program has a maximum memory usage of less than 2 GB (cf.
the table in Figure \ref{tab:full_RDP_main_results}) when clustering the
\texttt{RDP} dataset of size 3.5 GB (cf. section
\ref{sec:overview_of_datasets}), so storing all data in memory is not actually
a problem for the sizes of the datasets used for testing.


\subsection{Speeding up centroid search}

For each sequence, a \verb|bitset<4096>|, i.e. a space efficient \texttt{C++}
bit vector of statically fixed size, of the $6$-mers occurring in that
sequence, is computed. If that sequence becomes a centroid, the bit vector is
copied and stored in a structure representing the centroid, along with a
reference to the centroid sequence etc., to avoid recomputing it later. If the
sequence matches an existing centroid, the bit vector is reset and reused for
subsequent sequences. This yields a memory usage of 4096 bits times the number
of centroids, in addition to the sequence data.

Having bit vectors of the $k$-mers occurring in both a sequence and a centroid
sequence allows for very efficient computation of a bit vector of the $k$-mers
occurring in both the sequence and the centroid sequence. This is done using
the bitwise \verb|AND| operation. Additionally, computing the number of set
bits in a \verb|bitset| is extremely efficient due to \emph{single instruction,
multiple data} (SIMD) instructions on newer Intel Core processors~\cite{intel}
(beginning with the Nehalem microarchitecture) which are enabled when compiling
with the \verb|-march=native| compiler flag using the GCC \texttt{C++}
compiler.


\subsection{Divide and conquer with parallelization of clustering}

A simple divide and conquer approach was implemented and used with the
\textsc{K-Clust} algorithm. This is, however, not used by default in the
\texttt{klust} program, although it can be enabled with a command line flag.

The array of sequences is divided into a given number of subarrays, using
indexing to avoid copying, which are then each clustered in separate threads
using the implementation of \textsc{K-Clust}.  Each of these clustering
subtasks yields a list of centroids which are then merged in a similar way to
how \textsc{K-Clust} searches for a centroid: two centroids representing
different clusters are compared if they satisfy the intersection criterion.
Additionally, $max\_rejects$ is also used to limit the number of compares that
can be made for each centroid.
